---
title: "Exploratory Data Analysis (Disaster-Agnostic)"
format: pdf
---

## Instructions

Make sure you set `tzone` in the code chunk below to the relevant timezone. Then, you can run this file to get rough population and displacement plots for the first and last timepoints.

## Data Cleaning + Loading
```{r}
re_run_cleaning <- T
tzone <- #### ADD TIMEZONE HERE
source("1_data_cleaning.R")
source("2_plot_functions.R")
```


## First Timestamp

### Population Map

```{r}
population_plot_n_crisis(plot_ds = w[["first_ds"]],
                             plot_hour = w[["first_hour"]],
                             lon_limits = range(fb_data_bing$longitude), 
                             lat_limits = range(fb_data_bing$latitude),
                             zoom = 6,
                             tzone = tzone
                             )

```

### Displacement Map

```{r}
population_plot_n_difference(plot_ds = w[["first_ds"]],
                             plot_hour = w[["first_hour"]],
                             lon_limits = range(fb_data_bing$longitude), 
                             lat_limits = range(fb_data_bing$latitude),
                             zoom = 6,
                             tzone = tzone
                             )

```

## Last Timestamp

### Population Map

```{r}
population_plot_n_crisis(plot_ds = w[["latest_ds"]],
                             plot_hour = w[["latest_hour"]],
                             lon_limits = range(fb_data_bing$longitude), 
                             lat_limits = range(fb_data_bing$latitude),
                             zoom = 6,
                             tzone = tzone
                             )

```

### Displacement

```{r}
population_plot_n_difference(plot_ds = w[["latest_ds"]],
                             plot_hour = w[["latest_hour"]],
                             lon_limits = range(fb_data_bing$longitude), 
                             lat_limits = range(fb_data_bing$latitude),
                             zoom = 6,
                             tzone = tzone
)

```

## Animations

```{r}
source("5_animations.R")
```

### Population Animation

```{r}
population_plot_animate(
  metric = "crisis",
  tzone = tzone,
  lon_limits = range(fb_data_bing$longitude),
  lat_limits = range(fb_data_bing$latitude),
  fps = 8,
  filename = "crisis.gif"
)

```

### Population Animation

```{r}
population_plot_animate(
  metric = "difference",
  tzone = tzone,
  lon_limits = range(fb_data_bing$longitude),
  lat_limits = range(fb_data_bing$latitude),
  fps = 8,
  filename = "difference.gif"
)
```
