---
title: "Nepal Glacier Outburst Flood: Facebook Data Situation Report"
subtitle: |
  [NSF AI-SDM](https://www.cmu.edu/ai-sdm/) and Meta AI for Good\
  Report Date: October 7, 2026 · Latest Data: September 8, 2026
author:
  - Meta Disaster Analysts            # authors
format:
  typst:
    papersize: us-letter
    margin: { x: 1.5cm, y: 1.5cm }
    fontsize: 10pt
    keep-typ: true
    fig-format: png
    fig-dpi: 300
execute:
  echo: false
  warning: false
  message: false
  output: true
---

# Disaster Manager Brief - Nepal Glacier Outburst Flood

```{r setup}
library(tidyverse)
library(sf)
library(lubridate)
library(ggforce)
library(ggspatial)
library(maptiles)
library(quadkeyr)
library(prettymapr)
library(patchwork)
library(terra)      # used by 3_flood_functions.R (terrain)
library(elevatr)
library(osmdata)

# ---- Data + plotting functions (same scripts you already use) ---------------
# NOTE: 1_data_cleaning.R sets re_run_cleaning <- TRUE and calls
# load_population(re_run = TRUE) itself, so every render re-reads the raw CSVs.
# For faster renders, change those two spots to use the cached .Rds files.
env <- new.env()
env$re_run_cleaning <- FALSE
source("1_data_cleaning.R", local = env)

# 2_plot_functions.R: load ONLY the function definitions. Everything from the
# "Define coordinate limits..." comment onward is demo plotting code that would
# otherwise print stray figures into the report.
src <- readLines("2_plot_functions.R")
cut <- grep("Define coordinate limits directly from your dataset", src, fixed = TRUE)[1]
stopifnot(!is.na(cut))
eval(parse(text = src[seq_len(cut - 1)]), envir = env)

attach(env)   # makes w, first_ds, latest_ds, tzone, fb_data_bing, ... visible
```

```{r setup-data}
flood_date_txt <- "26 August 2026"          # date of the glacier outburst

# Map box (extended north so the outburst site is inside, and west for labels)
lon_limits <- range(fb_data_bing$longitude, na.rm = TRUE)
lat_limits <- range(fb_data_bing$latitude,  na.rm = TRUE)
lat_limits[2] <- max(lat_limits[2], 28.45)
lon_limits[1] <- lon_limits[1] - 0.25

# All available windows, and three evenly spaced ones (first / middle / latest)
win  <- fb_data_bing |>
  distinct(ds, hour) |>
  mutate(ds = as.character(ds)) |>
  arrange(ds, hour)
pick <- unique(round(seq(1, nrow(win), length.out = 3)))
sel  <- paste(win$ds[pick], win$hour[pick])

# One symmetric colour scale shared by every difference map in the report
.m <- fb_data_bing |>
  filter(paste(as.character(ds), hour) %in% sel) |>
  pull(n_difference) |>
  abs() |>
  max(na.rm = TRUE)
.dl <- c(-.m, .m)

# Net change in users inside a small box around a place, for one window
site_summary <- function(lon, lat, ds_ = first_ds, hr = first_hour) {
  d <- fb_data_bing |>
    filter(as.character(ds) == ds_, hour == hr,
           between(longitude, lon[1], lon[2]),
           between(latitude,  lat[1], lat[2]),
           !is.na(n_difference))
  list(net = sum(d$n_difference), tiles = nrow(d), area = 2.4 * 2.4 * nrow(d))
}

# Boxes around each place (~20 km); adjust if you want a tighter / wider window
places <- tribble(
  ~place,        ~role,                     ~lon,      ~lat,
  "Glacier outburst site", "Outburst origin", 85.5000,  28.3000,
  "Kathmandu",   "Capital, valley floor",   85.3240,   27.7172,
  "Bharatpur",   "Chitwan, south-west",     84.4333,   27.6833,
  "Hetauda",     "Makwanpur, south",        85.029716, 27.429071
)
half <- 0.1   # degrees either side of each place
sums <- purrr::pmap(places, function(place, role, lon, lat, ...)
  site_summary(lon + c(-half, half), lat + c(-half, half)))
places$net   <- purrr::map_dbl(sums, "net")
places$tiles <- purrr::map_int(sums, "tiles")
places$area  <- purrr::map_dbl(sums, "area")

fmt <- function(x) prettyNum(round(abs(x), 0), big.mark = ",")
ktm  <- places |> filter(place == "Kathmandu")
bha  <- places |> filter(place == "Bharatpur")
het  <- places |> filter(place == "Hetauda")
dir_word <- function(x) ifelse(x >= 0, "more", "fewer")
mov_word <- function(x) ifelse(x >= 0, "entered", "exited")

# Shared plotting helpers ------------------------------------------------------
flood_panel <- function(ds, hr, metric = c("difference", "crisis")) {
  metric <- match.arg(metric)
  f <- if (metric == "difference") population_plot_n_difference else population_plot_n_crisis
  args <- list(plot_ds = ds, plot_hour = hr, tzone = tzone,
               lon_limits = lon_limits, lat_limits = lat_limits, zoom = 6,
               highway_detail = "none", disaster_type = "flood")
  if (metric == "difference") args$fill_limits <- .dl
  do.call(f, args)
}

legend_guides <- ggplot2::guides(
  fill   = ggplot2::guide_colorbar(title.position = "top", order = 1,
                                   barwidth = grid::unit(6, "cm"),
                                   barheight = grid::unit(0.35, "cm")),
  colour = ggplot2::guide_legend(title.position = "left", order = 2, nrow = 1,
                                 keywidth = grid::unit(1, "cm")))
legend_theme <- ggplot2::theme(
  legend.position = "bottom", legend.box = "vertical",
  legend.text = ggplot2::element_text(size = 7),
  legend.title = ggplot2::element_text(size = 8),
  plot.margin = ggplot2::margin(2, 6, 2, 6))
```

## Key Takeaways

* **Outburst and first window.** A glacier outburst flood began near the glacier site (about 28.3°N, 85.5°E) on **`r flood_date_txt`**. The first available data window is `r first_label`; the latest is `r latest_label` (@fig-diff).

* **Kathmandu.** In the first window, `r fmt(ktm$net)` `r dir_word(ktm$net)` Facebook users than the prior 90-day baseline were present in the Kathmandu valley box (roughly `r round(ktm$area)` square kilometers; @fig-diff, @fig-ts).

* **Bharatpur.** `r fmt(bha$net)` `r dir_word(bha$net)` users than baseline in the first window (roughly `r round(bha$area)` square kilometers).

* **Hetauda.** `r fmt(het$net)` `r dir_word(het$net)` users than baseline in the first window (roughly `r round(het$area)` square kilometers).

* **Persistence.** The time series (@fig-ts) shows how the net change evolves through `r latest_label`; compare the first and latest maps (@fig-diff) to see how far activity has moved back toward baseline.

* Differences are relative to a 90-day baseline and do **not** imply large or small raw user counts (@fig-counts).

```{r}
#| label: tbl-places
#| tbl-cap: "Locations used in this report, with the net change in users (crisis − baseline) in a ~20 km box around each, in the first window."
places |>
  transmute(Place = place,
            Role = role,
            `Latitude (°N)` = lat,
            `Longitude (°E)` = lon,
            `Net change in users` = round(net),
            `Tiles with data` = tiles) |>
  knitr::kable(digits = 4, format.args = list(big.mark = ","))
```

## Overview

This report describes population movements around the **`r flood_date_txt`** glacier outburst flood in Nepal, using data from Meta on *Facebook Population During Crisis* from `r first_label` to `r latest_label`. Each map compares the number of Facebook users in an area during the crisis window with a prior 90-day baseline for the same time of day and day of week (@tbl-places lists the places discussed).

1. **Kathmandu valley.** The map of the first window (the day of the outburst) shows the pattern of user influx and exodus around Kathmandu (@fig-diff), and the time series shows how it changes through the later windows (@fig-ts).

2. **Surrounding towns.** Bharatpur (west) and Hetauda (south) are marked on every map for context; their net change is in @tbl-places.

3. Population differences (relative to a 90-day baseline) do not imply large or small raw counts present in a city. Raw counts can remain high throughout the crisis (@fig-counts).

> **Before circulating:** the numbers above are computed directly from the data. Please read them against the maps and edit the interpretive sentences (e.g. whether a pattern reflects movement vs. a connectivity loss; see the caveats in the Appendix).

## 1. Per-tile user population: Difference in crisis window vs. pre-crisis baseline

The maps plot (crisis − baseline) per tile. **Red = fewer users than baseline (decrease); dark green = more users than baseline (increase); white = no change; grey = privacy threshold (NA).** Contours (every 500 m; bolder every 1,000 m), major rivers and major roads (OpenStreetMap) are overlaid; the orange × marks the glacier outburst site. Where cells are shaded strongly they are drawn above the terrain and rivers.

```{r}
#| label: fig-diff
#| fig-width: 7.3
#| fig-height: 5.4
#| out-width: "100%"
#| fig-cap: "Difference (crisis − baseline) per tile. Left = first window, right = latest window. Shared symmetric colour scale. Red = fewer than baseline, dark green = more than baseline, grey = NA."
options(flood_panel_width_in = 3.2, flood_label_size = 2.4)
((flood_panel(first_ds, first_hour) | flood_panel(latest_ds, latest_hour)) +
    patchwork::plot_layout(guides = "collect")) & legend_guides & legend_theme
```

#### Evolution through the crisis

```{r}
#| label: fig-seq
#| fig-width: 7.3
#| fig-height: 4.2
#| out-width: "100%"
#| fig-cap: "The same map at three evenly spaced windows (first, middle, latest). Same colour scale as the figure above."
options(flood_panel_width_in = 2.1, flood_label_size = 1.8)
panels <- map2(win$ds[pick], win$hour[pick], ~ flood_panel(.x, .y))
(wrap_plots(panels, nrow = 1) + patchwork::plot_layout(guides = "collect")) & legend_guides & legend_theme
```

```{r}
#| label: fig-ts
#| fig-width: 7.3
#| fig-height: 3.2
#| out-width: "100%"
#| fig-cap: "Net change in users (crisis − baseline), summed over a ~20 km box around each place, by data window. The dashed orange line marks the outburst (26 August, 08:37 Nepal time); times are Nepal time (Facebook windows are stamped in Pacific Time and converted). Positive = more users than baseline."
ts <- map_dfr(seq_len(nrow(places)), function(i) {
  p <- places[i, ]
  if (p$place == "Glacier outburst site") return(NULL)   # no population there
  fb_data_bing |>
    filter(between(longitude, p$lon - half, p$lon + half),
           between(latitude,  p$lat - half, p$lat + half)) |>
    group_by(ds, hour) |>
    summarise(net = sum(n_difference, na.rm = TRUE), .groups = "drop") |>
    mutate(place = p$place,
           t = ymd_h(paste(as.character(ds), substr(hour, 1, 2)),
                     tz = "America/Los_Angeles") |>
             with_tz("Asia/Kathmandu"))
})
ggplot(ts, aes(t, net)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_vline(xintercept = ymd_hm("2026-08-26 08:37", tz = "Asia/Kathmandu"),
             linetype = "dashed", colour = "darkorange") +
  geom_line(colour = "#1A2421", linewidth = 0.5) +
  geom_point(colour = "#1A2421", size = 0.8) +
  facet_wrap(~ place, nrow = 1, scales = "free_y") +
  labs(x = "Window start (Nepal time)", y = "Net users vs. baseline") +
  scale_y_continuous(labels = scales::label_comma()) +
  theme_minimal(base_size = 8) +
  theme(panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 2. Raw user counts

Raw users per tile in the latest window show where users **are**, not where change occurred. Metro areas (Kathmandu valley in particular) remain the most heavily used even when they differ from baseline.

```{r}
#| label: fig-counts
#| fig-width: 7.3
#| fig-height: 5
#| out-width: "70%"
#| fig-cap: "Raw crisis user counts per tile in the latest window (log scale)."
options(flood_panel_width_in = 3.2, flood_label_size = 2.4)
flood_panel(latest_ds, latest_hour, metric = "crisis") &
  ggplot2::theme(legend.position = "bottom")
```

{{< pagebreak >}}

## Appendix

## Data description

We use Meta's **Facebook Population During Crisis** dataset, which displays location activity from Facebook users (with **Location Services on**) to estimate how population shifts during a crisis. The data represent **Facebook app users**, not everyone in the area, henceforth "users". The dataset compares user behavior near the incident with pre-crisis behavior at two levels: fixed **8-hour windows** and **Bing tiles** (level 14, ~2.4 km or ~1.5 mi squares). For privacy, a window is shown as NA / grey when either the crisis **or** the baseline count is under 10 users. Grey indicates "one side was below threshold" (not "no change" or "fewer than 10 people total"). Two signals are used: comparison to a **pre-crisis baseline** for the same time-of-day / day-of-week (a 90-day baseline), and **raw user counts** at window *T*.

As of this report the latest **population** window is `r latest_label`. The maps in Section 1–2 therefore show windows from `r first_label` to `r latest_label`.

::: {.callout-warning title="Data caveats" appearance="simple"}
- **Sample bias.** This data covers only Facebook users with Location Services turned on.
- **Data intervals.** 8-hour windows, up to ~8 h delay.
- **Tile size.** ~2.4 km or ~1.5 mi, neighborhood-scale, not address-level.
- **Privacy controls.** A window is NA / grey when the crisis or baseline count is <10 (both <10 means the row is dropped).
- **Baselines.** Population data use a 90-day baseline before the crisis.
- **Evacuation vs connectivity.** A tile emptying could mean people left **or** power / cell service dropped, which is plausible in flood-affected, mountainous terrain.
- **Terrain.** Valley and mountain geography can concentrate both movement and signal loss in a few tiles; read the contour lines when comparing areas.
- **Who is under-counted.** The signal skews toward mobile, connected, vehicle-owning residents and may under-capture access-and-functional-needs populations, such as people with mobility limits, no vehicle, or no transit option.
- **Users ≠ people ≠ vehicles.** Counts are a fraction of residents. These figures are not a road-traffic count.
:::

#### Full scope of data

```{r}
#| label: fig-full
#| fig-width: 7.3
#| fig-height: 5.4
#| out-width: "100%"
#| fig-cap: "Overview map of the entire data extent used in this report: user locations (blue) with Kathmandu and the glacier outburst site marked."
loc <- fb_data_bing |> distinct(longitude, latitude)
ggplot(loc, aes(longitude, latitude)) +
  geom_point(colour = "blue", alpha = 0.15, size = 0.4) +
  annotate("point", x = 85.3240, y = 27.7172, colour = "red", shape = 4, stroke = 1.2, size = 4) +
  annotate("label", x = 85.3240 + 0.12, y = 27.7172 - 0.1, label = "Kathmandu",
           fontface = "bold", fill = "white", size = 3) +
  annotate("point", x = 85.5, y = 28.3, colour = "darkorange", shape = 4, stroke = 1.5, size = 4) +
  annotate("label", x = 85.5 - 0.3, y = 28.3, label = "Glacier outburst site",
           fontface = "bold", fill = "white", size = 3) +
  coord_fixed() +
  labs(x = "Longitude (°E)", y = "Latitude (°N)") +
  theme_minimal(base_size = 8)
```

#### Sources

* Meta : Facebook Population During Crisis.
* Map data © OpenStreetMap contributors (rivers, roads); terrain from AWS Terrain Tiles via the `elevatr` package; basemap tiles via `maptiles`.
